### 2013-06-03 doc/dissertation/document/chapters/mc_background/mc_background.tex
updating spn chapter

\subsubsection{Alternative Fixed Point Iterations}
\label{subsubsec:alternative_fixed_point}
In addition to the basic Richardson iteration, the MCSA algorithm
presented in Eq~(\ref{eq:mcsa}) can be used to accelerate any fixed
point iteration which only depends on the state (typically the
residual) of the last iteration. As outlined in
Appendix~\ref{ch:linear_problem}, subspace methods take on a general
form in one-dimension using the Petrov-Galerkin conditions as
constraints for extracting a correction from the search subspace as
given by Eq~(\ref{eq:linear_projection_iteration}) that may be used
with MCSA. As an example, for positive-definite but not necessarily
symmetric problems, the minimal residual (MINRES) iteration
\citep{saad_iterative_2003} can be used in conjuction with MCSA where
the residual vector, $\mathbf{r}$, defines the search subspace and the
action of the linear operator on the residual vector,
$\mathbf{A}\mathbf{r}$, defines the constraint subspace. Using this,
we can then define an MCSA scheme that accelerates the MINRES
iteration:
\begin{subequations}
  \begin{gather}
    \alpha = \frac{\langle \mathbf{A}\mathbf{r}^k, \mathbf{r}^k
      \rangle}{\langle \mathbf{A}\mathbf{r}^k, \mathbf{A}\mathbf{r}^k
      \rangle}\:, \\
    \ve{x}^{k+1/2} = \ve{x}^k + \alpha \ve{r}^k\:,
    \label{eq:minres_correction}\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\ 
    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\ 
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta\ve{x}^{k+1/2}\:,\\ 
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:.
  \end{gather}
  \label{eq:mcsa_minres}
\end{subequations}
where $\alpha$ is an optimal extrapolation parameter generated the
constraints. Interestingly, this scheme is nearly identical to the
Richardson iteration version in Eq~(\ref{eq:mcsa}) except now the
additional extrapolation parameter, $\alpha$, is applied to the
residual correction in Eq~(\ref{eq:minres_correction}) as a means of
selecting a more optimal search direction at each iteration,
potentially further accelerating convergence.

@@

\subsection{Reduced Domain Approximation}
\label{subsec:reduced_domain_approximation}
For scalable parallel implementations and efficient Monte Carlo
transport, the domain is required to be sparse. We may find through
general preconditioning operations as outlined by the Monte Carlo
problem in Eq~(\ref{eq:left_right_correction}) that this sparsity is
lost. To alleviate this, we propose modifying the domain sampled by
the Monte Carlo method to generate the correction within MCSA by
removing terms in the iteration matrix through some predetermined
criteria in order to recover sparsity. As the Neumann-Ulam scheme is a
stochastic realization of multiple matrix-vector multiplies, the
criteria we choose should be based on maintaining those elements that
will make the largest contributions to the multiplication and
therefore recover as much of the original product as possible. For
each row in the preconditioned system, we can then apply the reduced
domain approximation to eliminate those elements that either fall
below a specified tolerance level, keep the $N$ largest elements in
each row, or both giving:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
    \hat{\ve{A}}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:, 
    \label{eq:rda_correction}\\
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
  \end{gather}
  \label{eq:rda_mcsa}
\end{subequations}
where $\hat{\ve{A}}$ in Eq~(\ref{eq:rda_correction}) means that the
reduced domain approximation has been applied to the iteration matrix
in the Monte Carlo method. If enough of the terms in the original
iteration matrix have been maintained, the character of the original
iteration matrix should be retained and continue to accelerate the
iterative scheme. The effects of the reduced domain approximation on
systems where it is required due to the explicit preconditioning
strategy chosen will be presented in Chapter~\ref{ch:spn_equations}.

>>>>

\subsubsection{Alternative Fixed Point Iterations}
\label{subsubsec:alternative_fixed_point}
In addition to the basic Richardson iteration, the MCSA algorithm
presented in Eq~(\ref{eq:mcsa}) can be used to accelerate any fixed
point iteration which only depends on the state (typically the
residual) of the last iteration. As outlined in
Appendix~\ref{ch:linear_problem}, subspace methods take on a general
form in one-dimension using the Petrov-Galerkin conditions as
constraints for extracting a correction from the search subspace as
given by Eq~(\ref{eq:linear_projection_iteration}) that may be used
with MCSA. As an example, for positive-definite but not necessarily
symmetric problems, the minimal residual (MINRES) iteration
\citep{saad_iterative_2003} can be used in conjuction with MCSA where
the residual vector, $\mathbf{r}$, defines the search subspace and the
action of the linear operator on the residual vector,
$\mathbf{A}\mathbf{r}$, defines the constraint subspace. Using this,
we can then define an MCSA scheme that accelerates the MINRES
iteration:
\begin{subequations}
  \begin{gather}
    \alpha = \frac{\langle \mathbf{A}\mathbf{r}^k, \mathbf{r}^k
      \rangle}{\langle \mathbf{A}\mathbf{r}^k, \mathbf{A}\mathbf{r}^k
      \rangle}\:, \\
    \ve{x}^{k+1/2} = \ve{x}^k + \alpha \ve{r}^k\:,
    \label{eq:minres_correction}\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\ 
    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\ 
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta\ve{x}^{k+1/2}\:,\\ 
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:.
  \end{gather}
  \label{eq:mcsa_minres}
\end{subequations}
where $\alpha$ is an optimal extrapolation parameter generated from
the constraints. Interestingly, this scheme is nearly identical to the
Richardson iteration version in Eq~(\ref{eq:mcsa}) except now the
additional extrapolation parameter, $\alpha$, is applied to the
residual correction in Eq~(\ref{eq:minres_correction}) as a means of
selecting a more optimal search direction at each iteration,
potentially further accelerating convergence.

@@

\subsection{Reduced Domain Approximation}
\label{subsec:reduced_domain_approximation}
For scalable parallel implementations and efficient Monte Carlo
transport, the domain is required to be sparse. We may find through
general preconditioning operations as outlined by the Monte Carlo
problem in Eq~(\ref{eq:left_right_correction}) that this sparsity is
lost. To alleviate this, we propose modifying the domain sampled by
the Monte Carlo method to generate the correction within MCSA by
removing terms in the iteration matrix through some predetermined
criteria in order to recover sparsity. As the Neumann-Ulam scheme is a
stochastic realization of multiple matrix-vector multiplies, the
criteria we choose should be based on maintaining those elements that
will make the largest contributions to the multiplication and
therefore recover as much of the original product as possible. For
each row in the preconditioned system, we can then apply the reduced
domain approximation to eliminate those elements that either fall
below a specified tolerance level, keep the $N$ largest elements in
each row, or both giving:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
    \hat{\ve{A}}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:, 
    \label{eq:rda_correction}\\
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
  \end{gather}
  \label{eq:rda_mcsa}
\end{subequations}
where $\hat{\ve{A}}$ in Eq~(\ref{eq:rda_correction}) means that the
reduced domain approximation has been applied to the iteration matrix
in the Monte Carlo method. If enough of the terms in the original
iteration matrix have been maintained, the character of the original
iteration matrix should be retained and continue to accelerate the
iterative scheme. The effects of the reduced domain approximation on
neutron transport systems where it is required due to the explicit
preconditioning strategy will be presented in
Chapter~\ref{ch:spn_equations}.

### 2013-06-03 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updating spn chapter

are provided in Appendix~\ref{chap:spn_boundary_conditions}. The
multigroup form is presented in Appendix~\ref{chap:spn_mg_equations}

@@

\subsection{MCSA Verification}
\label{subsec:spn_mcsa_verification}
In order to verify the correctness of the MCSA method and
implementation used for this work, MCSA solutions to the fuel assembly
problem with 1, 2, and 4 energy groups using $SP_1$ discretization
will be directly compared over all group and cell fluxes computed by
converging the criticality problem using both BiCGStab and GMRES
production implementations from the Trilinos scientific computing
library Aztec \citep{heroux_overview_2005}. In addition to comparing
the cell and group fluxes, the number of eigenvalue iterations
required to converge the problem along with the eigenvalue itself will
be compared as a means of verification. For the verification, all
solvers will be preconditioned with ILUT using a drop tolerance of
\sn{1}{-5} and a fill level of 5. For GMRES, no restrictions were
placed on the size of the subspace and therefore no restarts
occurred. When the collision estimator was used with MCSA, 20,000
stochastic histories were used to compute the correction for every
energy group in the problem (40,000 and 80,000 histories total for the
2 and 4 group calculations respectively) and 500 stochastic histories
used for each energy group when the expected value estimator was used
(1,000 and 1,500 histories total for the 2 and 4 group
calculations). The reduced domain approximation was also applied in
conjuction with the ILUT preconditioning using a fill level of 100 and
a threshold value of \sn{1}{-10} to reduce the density of states in
the Monte Carlo problem. For relaxation parameters, all MCSA
compuations used a Neumann relaxation parameter of 0.7 while a
Richardson relaxation of 1.1 was used with the collision estimator and
1.0 with the expected value estimator as determined by the previous
analysis. Table~\ref{tab:spn_solver_defs} gives definitions for the
solvers used to generate the results in the remainder of this section.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{l}{\textbf{Name}} & 
      \multicolumn{1}{l}{\textbf{Definition}} \\
      BiCGStab-ILUT & ILUT preconditioned BiCGStab \\
      GMRES-ILUT & ILUT preconditioned GMRES \\
      MCSA-ILUT-R-C & ILUT preconditioned MCSA with Richardson \\ 
                    & fixed point iteration and collision estimator \\
      MCSA-ILUT-MR-C & ILUT preconditioned MCSA with MINRES \\
                     & fixed point iteration and collision estimator \\
      MCSA-ILUT-R-EV & ILUT preconditioned MCSA with Richardson \\
                     & fixed point iteration and expected value estimator \\
      MCSA-ILUT-MR-EV & ILUT preconditioned MCSA with MINRES \\
                      & fixed point iteration and expected value estimator \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Solver definitions used for MCSA verification and
      performance analysis.}}
  \label{tab:spn_solver_defs}
\end{table}
For the energy group structures,
Table~\ref{tab:spn_group_structure} gives the lower bounds of each
group in eV (an implicit upper bound of 20 MeV is assumed).
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cl}\hline\hline
      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
      \multicolumn{1}{l}{\textbf{Lower Bounds (eV)}} \\
      1 & \{ \sn{1}{-5} \} \\
      2 & \{ \sn{1}{-1}, \sn{1}{-5} \} \\
      4 & \{ \sn{1}{1}, \sn{1}{0}, \sn{1}{-1}, \sn{1}{-5} \} \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Energy group lower bounds for the multigroup fuel
      assembly criticality problem in electron volts.}}
  \label{tab:spn_group_structure}
\end{table}

@@

Figure~\ref{fig:spn_comparison_iterations} gives the average number of
linear solver iterations required to converge a single eigenvalue
iteration of the fuel assembly problem as a function of the number of
energy groups. In general, the iterative performance of all the
solvers tested is comparable with BiCGStab performing the best. We
expect this because not only are the multigroup $SP_N$ equations
elliptic and positive-definite, but they are also nearly symmetric and
therefore we expect the conjugate gradient-based methods to perform
well. It also important to note that the iterative performance of each
solver is not a strong function of the number of energy groups in the
problem (and therefore the problem size). For MCSA, we are linearly
increasing the number of stochastic histories used to compute the
correction at each iteration as the number of energy groups are
increased and therefore the work performed for the problem is of
$O(N)$. It is also important to note the the reduced domain
approximation parameters, particularly the fill level, were fixed as
the number of energy groups was increased. By fixing the fill level,
we are effectively fixing the amount of information contained in the
composite linear operator available for the Monte Carlo
problem. Limiting this information to 100 entries per row in the
preconditioned system did not appear to have a significant effect on
the iterative performance of the solver. For larger numbers of energy
groups (and therefore DOFs) this may be an issue and this number may
need to be increased to maintain iterative performance. Currently,
memory restrictions arising from the explicit preconditioning strategy
prevent larger energy groups to be analyzed using the appropriate
preconditioning for the fuel assembly problem.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/solver_iters.pdf}
  \end{center}
  \caption{\textbf{Average number of iterations required to converge
      each eigenvalue iteration for the fuel assembly problem as a
      function of energy groups.} \textit{All methods were
      preconditioned with identical ILUT parameters. For MCSA, several
      variations of fixed point iterations and estimators were used.}}
  \label{fig:spn_comparison_iterations}
\end{figure}

Although iterative performance for MCSA was comparable to that
observed for production Krylov implementations using exactly the same
preconditioning scheme, timing performance is not as
comparable. Figure~\ref{fig:spn_comparison_time} gives the average CPU
time per linear solver iteration required to converge the fuel
assembly problem with all linear solver iterations over all eigenvalue
iterations used to compute the average. The production Krylov solver
implementations are approximately an order of magnitude faster than
the MCSA implementations presented here. We expect these results for
several reasons. First, even when the reduced domain approximation is
used there are O(100) elements in each row of the system and each of
those elements must be processed during a transition in the Monte
Carlo random walk sequence. In addition, although the preconditioned
fixed point iterations used in the MCSA sequence do not use the
composite linear operator but rather a sequence of matrix vector
multiplies to achieve the same effect, they do use the explicitly
extracted inverses of the preconditioning matrices which themselves
are dense, leading to exceedingly slow computation times even in the
fixed point iteration. Second, the implementation presented here has
not been optimized and it is likely several components of the
algorithm can be implemented in a more desireable time complexity. The
production Krylov solvers presented here have enjoyed decades of
professional development and optimization. However, it is most
important to note here that the CPU timing data presented here shows
that all methods have the same time complexity; their differences are
merely the manifestation of a large time constant due to the effects
of the explicit preconditioning scheme. For transport systems where
this is not a problem, the literature explcitly shows that MCSA can be
competetive with these production Krylov methods using CPU time as a
measure \citep{evans_monte_2012}.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/solver_time.pdf}
  \end{center}
  \caption{\textbf{Average CPU time per linear solver iteration in
      seconds for the fuel assembly problem as a function of energy
      groups.}  \textit{All linear solver iterations over all
      eigenvalue iterations were used to compute the averge. All
      methods were preconditioned with identical ILUT parameters. For
      MCSA, several variations of fixed point iterations and
      estimators were used.}}
  \label{fig:spn_comparison_time}
\end{figure}

Finally, not considered in Figure~\ref{fig:spn_comparison_time} is the
time required to actually form the explicit inverse preconditioner
matrices and the composite operator through matrix-matrix
multiplication. The timing numbers reported were simply to perform the
CMSA iteration procedure with the composite operator already
formed. Figure~\ref{fig:spn_comparison_prec_time} additionally
presents the CPU times for MCSA convergence with the time to form the
preconditioners and the composite linear operator through
matrix-matrix multiplication amortized over all iterations. As is
readily observed including the costs of these operations increases the
MCSA computation time by another order of magnitude.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/solver_p_time.pdf}
  \end{center}
  \caption{\textbf{Average CPU time per iteration in seconds for the
      fuel assembly problem as a function of energy groups with
      preconditioning time included for the MCSA methods.}
    \textit{All linear solver iterations over all eigenvalue
      iterations were used to compute the averge. All methods were
      preconditioned with identical ILUT parameters. For MCSA, several
      variations of fixed point iterations and estimators were used.}}
  \label{fig:spn_comparison_prec_time}
\end{figure}

Based on the results in this section, MCSA shows promise as a
competitive and perhaps even superior method for solutions to the
neutron transport problem discretized with the $SP_N$
approximation. Not only are the correct answers produced when compared
to production linear solvers, but the iterative performance is
comparable to production Krylov methods with identical preconditioning
that would typically be used every-day calculations. From a CPU timing
perspective, the methods yield the same time complexity as a function
of the number of energy groups in the problem when compared to the
production Krylov methods. Here, a large time constant is generated
due to the explicit preconditioning strategy and the generation of
dense linear operators as a result. To improve these results and put
general MCSA schemes into a performance regime where they are
competitive with Krylov methods for neutron transport problems,
significant research will be required to improve upon the explicit
preconditioning shceme presented here.

>>>>

are provided in Appendix~\ref{chap:spn_boundary_conditions}, the
multigroup form is presented in Appendix~\ref{chap:mg_spn_equations},

@@

\subsection{MCSA Verification}
\label{subsec:spn_mcsa_verification}
In order to verify the correctness of the MCSA method and
implementation used for this work, MCSA solutions to the fuel assembly
problem with 1, 2, and 4 energy groups using $SP_1$ discretization
will be directly compared over all group and cell fluxes computed by
converging the criticality problem using both BiCGStab and GMRES
production implementations from the Trilinos scientific computing
library Aztec \citep{heroux_overview_2005}. In addition to comparing
the cell and group fluxes, the number of eigenvalue iterations
required to converge the problem along with the eigenvalue itself will
be compared as a means of verification.

For the verification, all solvers will be preconditioned with ILUT
using a drop tolerance of \sn{1}{-5} and a fill level of 5. For GMRES,
no restrictions were placed on the size of the subspace and therefore
no restarts occurred. When the collision estimator was used with MCSA,
20,000 stochastic histories were used to compute the correction for
every energy group in the problem (40,000 and 80,000 histories total
for the 2 and 4 group calculations respectively) and 500 stochastic
histories used for each energy group when the expected value estimator
was used (1,000 and 1,500 histories total for the 2 and 4 group
calculations). The reduced domain approximation was also applied in
conjuction with the ILUT preconditioning using a fill level of 100 and
a threshold value of \sn{1}{-10} to reduce the density of states in
the Monte Carlo problem. For relaxation parameters, all MCSA
compuations used a Neumann relaxation parameter of 0.7 while a
Richardson relaxation of 1.1 was used with the collision estimator and
1.0 with the expected value estimator as determined by the previous
analysis. Table~\ref{tab:spn_solver_defs} gives definitions for the
solvers used to generate the results in the remainder of this
section. For the energy group structures,
Table~\ref{tab:spn_group_structure} gives the lower bounds of each
group in eV (an implicit upper bound of 20 MeV is assumed).
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{l}{\textbf{Name}} & 
      \multicolumn{1}{l}{\textbf{Definition}} \\
      BiCGStab-ILUT & BiCGStab preconditioned with ILUT \\
      GMRES-ILUT & GMRES preconditioned with ILUT \\
      MCSA-ILUT-R-C & MCSA preconditioned with ILUT using Richardson \\ 
                    & fixed point iteration and collision estimator \\
      MCSA-ILUT-MR-C & MCSA preconditioned with ILUT using MINRES \\
                     & fixed point iteration and collision estimator \\
      MCSA-ILUT-R-EV & MCSA preconditioned with ILUT using Richardson \\
                     & fixed point iteration and expected value estimator \\
      MCSA-ILUT-MR-EV & MCSA preconditioned with ILUT using MINRES \\
                      & fixed point iteration and expected value estimator \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Solver definitions used for MCSA verification and
      performance analysis.}}
  \label{tab:spn_solver_defs}
\end{table}
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cl}\hline\hline
      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
      \multicolumn{1}{l}{\textbf{Lower Bounds (eV)}} \\
      1 & \{ \sn{1}{-5} \} \\
      2 & \{ \sn{1}{-1}, \sn{1}{-5} \} \\
      4 & \{ \sn{1}{1}, \sn{1}{0}, \sn{1}{-1}, \sn{1}{-5} \} \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Energy group lower bounds for the multigroup fuel
      assembly criticality problem in electron volts.}}
  \label{tab:spn_group_structure}
\end{table}

@@

Figure~\ref{fig:spn_comparison_iterations} gives the average number of
linear solver iterations required to converge a single eigenvalue
iteration of the fuel assembly problem as a function of the number of
energy groups. In general, the iterative performance of all the
solvers tested is comparable with BiCGStab performing the best. We
expect this because not only are the multigroup $SP_N$ equations
elliptic and positive-definite, but they are also nearly symmetric and
therefore we expect the conjugate gradient-based methods to perform
well. It also important to note that the iterative performance of each
solver is not a strong function of the number of energy groups in the
problem (and therefore the problem size). For MCSA, we are linearly
increasing the number of stochastic histories used to compute the
correction at each iteration as the number of energy groups are
increased and therefore the work performed for the problem is of
$O(N)$. It is also important to note the the reduced domain
approximation parameters, particularly the fill level, were fixed as
the number of energy groups was increased. By fixing the fill level,
we are effectively fixing the amount of information contained in the
composite linear operator available for the Monte Carlo
problem. Limiting this information to 100 entries per row in the
preconditioned system did not appear to have a significant effect on
the iterative performance of the solver. For larger numbers of energy
groups (and therefore DOFs) this may be an issue and this number may
need to be increased to maintain iterative performance. Currently,
memory restrictions arising from the explicit preconditioning strategy
prevent larger energy groups to be analyzed using an appropriate
preconditioner for the fuel assembly problem.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/solver_iters.pdf}
  \end{center}
  \caption{\textbf{Average number of iterations required to converge
      each eigenvalue iteration for the fuel assembly problem as a
      function of energy groups.}
    \textit{Table~\ref{tab:spn_solver_defs} gives the description for
      each solver type presented in the legend.}}
  \label{fig:spn_comparison_iterations}
\end{figure}

Although iterative performance for MCSA was comparable to that
observed for production Krylov implementations using exactly the same
preconditioning scheme, timing performance is not as
comparable. Figure~\ref{fig:spn_comparison_time} gives the average CPU
time per linear solver iteration required to converge the fuel
assembly problem with all linear solver iterations over all eigenvalue
iterations used to compute the average. The production Krylov solver
implementations are approximately an order of magnitude faster than
the MCSA implementations presented here. We expect these results for
several reasons. First, even when the reduced domain approximation is
used there are $O(100)$ elements in each row of the system and each of
those elements must be processed during a transition in the Monte
Carlo random walk sequence. In addition, although the preconditioned
fixed point iterations used in the MCSA sequence do not use the
composite linear operator but rather a sequence of matrix vector
multiplies to achieve the same preconditioning effect, they do use the
explicitly extracted inverse of the preconditioning matrices which
themselves are dense, leading to exceedingly slow computation times
even in the fixed point iteration. For the production Krylov methods,
the ILUT preconditioners are applied to a vector with two triangular
solves, one for each triangular factor. Also, the original linear
operator has $O(10)$ elements in each row of the system, an order of
magnitude less than that used in the reduced domain composite
operator. Second, the implementation presented here has not been
optimized and it is likely that several components of the algorithm
can be implemented in a more desireable time complexity. The
production Krylov solvers presented here have enjoyed decades of
professional development and optimization. However, it is most
important to note here that the CPU timing data presented in
Figure~\ref{fig:spn_comparison_time} shows that all methods have the
same time complexity; their differences are merely the manifestation
of a large time constant due to the effects of the explicit
preconditioning scheme. For transport systems where this is not a
problem, the literature explicitly shows that MCSA can be competetive
with these production Krylov methods using CPU time as a measure
\citep{evans_monte_2012}.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/solver_time.pdf}
  \end{center}
  \caption{\textbf{Average CPU time per linear solver iteration in
      seconds for the fuel assembly problem as a function of energy
      groups.}  \textit{All linear solver iterations over all
      eigenvalue iterations were used to compute the
      average. Table~\ref{tab:spn_solver_defs} gives the description
      for each solver type presented in the legend.}}
  \label{fig:spn_comparison_time}
\end{figure}

Finally, not considered in Figure~\ref{fig:spn_comparison_time} is the
time required to actually form the explicit inverse preconditioner
matrices and the composite operator through matrix-matrix
multiplication. The timing numbers reported were simply to perform the
MCSA iteration procedure with the composite operator already
formed. Figure~\ref{fig:spn_comparison_prec_time} additionally
presents the CPU times for MCSA convergence with the time to form the
inverse of the preconditioners and the composite linear operator
through matrix-matrix multiplication amortized over all iterations. As
is readily observed, including the costs of these operations increases
the MCSA computation time by another order of magnitude.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/solver_p_time.pdf}
  \end{center}
  \caption{\textbf{Average CPU time per iteration in seconds for the
      fuel assembly problem as a function of energy groups with
      preconditioning time included for the MCSA methods.}
    \textit{All linear solver iterations over all eigenvalue
      iterations were used to compute the
      average. Table~\ref{tab:spn_solver_defs} gives the description
      for each solver type presented in the legend. Data labeled
      starting with PMCSA is identical to those labeled with MCSA
      except that they additionally include the cost of generating the
      inverse of the preconditioners and composite linear operator.}}
  \label{fig:spn_comparison_prec_time}
\end{figure}

Based on the performance results in this section, MCSA shows promise
as a competitive and perhaps even superior method for solutions to the
neutron transport problem discretized with the $SP_N$
approximation. Not only are the correct answers produced when compared
to production linear solvers, but the iterative performance is
comparable to production Krylov methods with identical preconditioning
that would typically be used every-day calculations. From a CPU timing
perspective, the methods yield the same time complexity as a function
of the number of energy groups in the problem when compared to the
production Krylov methods. Here, a large time constant is generated
due to the explicit preconditioning strategy and the generation of
dense linear operators as a result. To improve these results and put
general MCSA schemes into a performance regime where they are
competitive with Krylov methods for neutron transport problems,
significant research will be required to improve upon the explicit
preconditioning scheme presented here.
